---
title: "Matrix Operations in R"
output: html_notebook
---

This notebook demonstrates how to do various matrix operations and matrix computations using R. 

In R, a matrix is defined as follows:
```{r}
A <- matrix(c(-1,0,2,-3,1,4),2,3,byrow = TRUE)
```

Let's look at the result:
```{r}
A
```

Observe,

  1) The entries of the matrix are provided by listing them with the concatenate function c()
  
  2) It is necessary to input the number of rows and number of columns
  
  3) If we want to matrix entries to be placed by row we must set byrow = TRUE
  
Once we have defined a matrix, the individual elements of the matrix are accessed as
```{r}
A[1,2]
```

or

```{r}
A[2,1]
```
and so on. 

R has a function the is designed to tell us the size or shape of a matrix:
```{r}
dim(A)
```


R makes it very easy to find the transpose of a matrix:
```{r}
t(A)
```

Let's look at how matrix multiplication works in R. 

Define a second matrix $B$ that we can multiply by $A$. 
```{r}
B <- matrix(c(1,3,-6,2,-1,0,1,-2,4,-2,7,-9,8,1,-3),3,5,byrow = TRUE)
```

Let's check the size of $B$ and then look at it. 
```{r}
dim(B)
```

```{r}
B
```

Notice that since $A$ is $2 \times 3$ and $B$ is $3 \times 5$ the product $AB$ is allowed and we expect the result to be a $2 \times 5$ matrix. We compute it as
```{r}
A %*% B
```

Let's use R to confirm the matrix identity $(AB)^{T} = B^T A^T$. 

```{r}
t(A %*% B)
```

```{r}
t(B) %*% t(A)
```


Now let's look at what we get when we compute $AA^T$ and $A^T A$.
```{r}
A %*% t(A)
```



```{r}
t(A) %*% A
```


Let's look at what happens if we create a matrix **without** setting byrow=TRUE:
```{r}
matrix(c(-1,0,2,-3,1,4),2,3)
```

This is sometimes conveniet say for example in order to create a column vector. 
```{r}
x <- matrix(c(-1,0,1),ncol=1)
```

```{r}
dim(x)
```

```{r}
x
```


Let's create another column vector with the same length as that of $x$. 
```{r}
y <- matrix(c(4,1,2),ncol = 1)
```

```{r}
y
```

Now we will compute $x^T y$
```{r}
t(x) %*% y
```

and also $x y^T$
```{r}
x %*% t(y)
```


There are simple functions that can be used to compute the determinant and inverse of a square matrix. 

```{r}
C <- matrix(c(-1,2,5,-1,0,2,2,1,-2),3,3,byrow = T)
```

```{r}
C
```

```{r}
det(C)
```

```{r}
solve(C)
```


Let's check that this is indeed the inverse to $C$.
```{r}
C_inv <- solve(C)
```

```{r}
C_inv %*% C
```


```{r}
C %*% C_inv
```

What do you observe about this? Let's see what happens if we round the result:
```{r}
round(C %*% C_inv,5)
```

Matrix addition is even easier to do than matrix multiplication:
```{r}
A <- matrix(c(1,2,3,4,5,6),2,3)
B <- matrix(c(2,3,4,5,6,7),2,3)
A + B
```


Finally, R can be used to solve systems of linear equations. 

Suppose we want to solve 

$-2x + 3y + z = 4$
$-x + 2y - z = -2$
$5x -2y + 3x = 8$

```{r}
A <- matrix(c(-2,3,1,-1,2,-1,5,-2,3),3,3,byrow = T)
b <- matrix(c(4,-2,8),ncol=1)
x <- solve(A,b)
```

Let's look at and then check the solution:
```{r}
x
```

```{r}
A %*% x
```

```{r}
b
```




